On Optimality of Finite Element Integration

### 2015-12-28 TOMS_Paper/assembly.tex
Minor fixes. Still need to ultimate the proof

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $T_m$ represent the set of tables resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$. If $T_{m_{1}} \cap T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary. $\Gamma$, therefore, may skip a potential global optimum, since $\theta$ only reasons on monomials in ``isolation''. We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a local -- instead of a global -- optimum. This situation is prevented by construction, because $\Gamma$ analyzes all possible bipartitions searching for a global optimum that also preserves the memory constraints. Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished. Finally, note how this problem can be seen as an instance of the well-known Knapsack problem.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Because of loop linearity and normal form of expressions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through (i) generalized code motion or (ii) factorization of multilinear symbols. These two strategies lead, respectively, to performing $|P|$ or $|S_P|$ products at every loop iteration. Since (i) is side effects free (i.e., neither expansion nor factorization are needed) and is applied if and only if $|P| < |S_P|$, it is clear that step (3) cannot prune any optimal solution from the search space; in fact, it just represents a move towards quasi-optimality.

Steps (4)-(5) factorize symbols following the ordering dictated by the sharing graph and its rewrite rules. Recall that rules are applied in sequence: first all rule (A) hits are handled, then all (B)s and finally all (C)s, with applications of rule (C) that may expose further possibilities for (A) and (B). We prove that the application of these rules, in this sequence, represents a step towards quasi-optimality. 
\begin{description}
\item[Rule (A)] $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule is side effects free (i.e., no need to choose between different factorizations).
\item[Rule (B)] Since $S^{nt}$ is no terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$. Let us consider the alternative factorization $S^{nt}(...S + ... + ...V)$. If $V$ were a terminal, we would simply be back to where we started, because rule (B) would be applicable by setting $V = T$ and $S = S^{nt}$. Otherwise ($V$ non terminal), we would have $S$ appearing in two different sub-expressions, which implies sub-optimality because of the presence of spatial sharing.
\item[Rule (C)] We can interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively. For each disconnected component $\mathbb{B^d} = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d) \in \mathbb{B}$ we consider two possibilities, namely $\mathbb{B^d}$ is 1) not complete or 2) complete. Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), the lower bound for the operation count (as number of products performed at every loop iteration) is given by ${min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$. This is reachable by taking the vertex with highest degree, as the rule states (note that other choices may break factorization opportunities). In case 2), the lower bound is 1 because factorization can be applied iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$); this is handled by going back to step (2), at the end of step (5).

% At this point, recall . 
%In order to have $\mathbb{T} = \emptyset$, which would trigger rule (C), it must be the case that ....
\end{description}
Finally, we note that step (6) is irrelevant for quasi-optimality.

\end{enumerate}

\end{proof}

@@

We evaluate the speed-ups achieved by a wide variety of transformation systems over the ``original'' code produced by the FEniCS Form Compiler (i.e., no optimizations applied). We analyze the following transformation systems
\begin{itemize}
\item FEniCS Form Compiler: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
\item FEniCS Form Compiler: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
\item FEniCS Form Compiler: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}). Referred to as \texttt{auto} 
\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation and execution times). Referred to as \texttt{ufls} 
\item COFFEE: generalized loop-invariant code motion (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
\item COFFEE: optimal loop nest synthesis plus symbolic execution for zero-elimination (work of this article). Referred to as \texttt{cfO2} 
\end{itemize}

>>>>

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $T_m$ represent the set of tables resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$. If $T_{m_{1}} \cap T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary. $\Gamma$, therefore, may skip a potential global optimum, since $\theta$ only reasons on monomials in ``isolation''. We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a local -- instead of a global -- optimum. This situation is prevented by construction, because $\Gamma$ analyzes all possible bipartitions searching for a global optimum that also preserves the memory constraints. Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished. Finally, note how this problem can be seen as an instance of the well-known Knapsack problem.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Because of loop linearity and normal form of expressions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through (i) generalized code motion or (ii) factorization of multilinear symbols. These two strategies lead, respectively, to performing $|P|$ or $|S_P|$ products at every loop iteration. Since (i) is side effects free (i.e., neither expansion nor factorization are needed) and is applied if and only if $|P| < |S_P|$, it is clear that step (3) cannot prune any optimal solution from the search space. %; in fact, it simply shifts towards the objective.

Steps (4)-(5) factorize symbols following the ordering dictated by the sharing graph and its rewrite rules. Recall that rules are applied in sequence: first all rule (A) hits are handled, then all (B)s and finally all (C)s, with applications of rule (C) that may expose further possibilities for (A) and (B). We prove that the application of these rules, in this sequence, represents a step towards quasi-optimality. 
\begin{description}
\item[Rule (A)] $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule is side effects free (i.e., no need to choose between different factorizations).
\item[Rule (B)] Since $S^{nt}$ is no terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$. Let us consider the alternative factorization $S^{nt}(...S + ... + ...V)$. If $V$ is a terminal, we are back to where we started, because rule (B) will be applicable by setting $V = T$ and $S = S^{nt}$. Otherwise ($V$ non terminal), we have $S$ appearing in two different sub-expressions, which implies sub-optimality because of the presence of spatial sharing.
\item[Rule (C)] We can interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively. For each disconnected component $\mathbb{B^d} = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$ in $\mathbb{G}$, we consider two possibilities, namely $\mathbb{B^d}$ 1) is not or 2) is complete. Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), the lower bound on the operation count (as number of products performed at every loop iteration), $lb$, is given by $lb = {min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$. This is reachable by factorizing the vertex with highest degree (as the rule states), as any other choice may break factorization opportunities. In case 2), we have $lb = 1$ because factorization is applicable iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$); this is handled by going back to step (2), at the end of step (5).
\end{description}
The algorithm is guaranteed to terminate because of the jump to step (6) when $\mathbb{S} = \emptyset$, before applying the rewrite rules. 
\end{enumerate}
\end{proof}

@@

We evaluate the speed-ups achieved by a wide variety of transformation systems over the ``original'' code produced by the FEniCS Form Compiler (i.e., no optimizations applied). We analyze the following transformation systems:
\begin{description}
\item[quad] Optimized quadrature mode. Work presented in~\citeN{quadrature1}, implemented in  in the FEniCS Form Compiler. 
\item[tens] Tensor contraction mode. Work presented in~\citeN{FFC-TC}, implemented in the FEniCS Form Compiler.
\item[auto] Automatic choice between \texttt{tens} and \texttt{quad} driven by heuristic (detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}). Implemented in the FEniCS Form Compiler.
\item[ufls] UFLACS, a novel back-end for the FEniCS Form Compiler whose primary goals are improved code generation and execution times.
\item[cfO1] Generalized loop-invariant code motion. Work presented in~\citeN{Luporini}, implemented in COFFEE.
\item[cfO2] Optimal loop nest synthesis with handling of block-sparse tables. Work presented in this article, implemented in COFFEE.
\end{description}
